Hydrous melting and partitioning in and above the mantle transition zone: insights from water-rich MgO-SiO$_2$-H$_2$O experiments

### 2015-07-12 MSH.tex
Update

\begin{abstract}

In this study, we conducted high pressure multianvil experiments at 13 GPa between 1200 and 1900 $^{\circ}$C to investigate the liquidus in the system MgO-SiO$_2$-H$_2$O. Water-rich starting compositions were obtained by using platinic acid (H$_2$Pt(OH)$_6$) as a novel water source. As MgO:SiO$_2$ ratios decrease, the liquidus curve develops an increasingly pronounced concave-down topology. At slab temperatures, the forsterite- and enstatite-water liquidii have similar water contents. Preliminary experiments on the SiO$_2$-H$_2$O binary reveal a pronounced melting point depression up to about 20 mol\% H$_2$O. This melting point depression exceeds estimates from published models, and requires either significant non-ideality between species in the melt, or distinct differences in the ease of protonation of bridging oxygens. We discuss the implications of this work on the migration of deep-seated slab fluids, and the partitioning of water between nominally anhydrous phases and melts.
\end{abstract}

@@

\begin{landscape}
\begin{table}[ht!]
\caption{Experimental run conditions and run products determined by SEM/EDS. Compositions are listed in order of increasing molar H$_2$O content. Unused compositions for each run are marked by an en-dash. Minor solids are listed in brackets. Mineral abbreviations are as follows: br - brucite, per - periclase, co - hydroxychondrodite, en - clinoenstatite, s - stishovite.}
\label{table:experiments}
\begin{tabular}{llllccccc}
Expt \# & P (GPa) & T ($^{\circ}$C) & t (min) & brucite+PtAc & forsterite+PtAcid & enstatite+PtAcid & quartz+PtAcid & br+q \\
Z1063 & 13 & 1200 & 40 & -/br/br/br/L & co/fo/fo/(fo) & en/en/en/en & -/-/-/-/-/-/-/- & -/-/-/- \\
Z1085 & 13 & 1300 & 40 & -/-/per/L/L & -/fo/(en)/L & -/-/-/- & -/-/-/-/-/-/-/- & -/-/-/- \\
Z1079 & 13 & 1300 & 35 & -/per/per/L/L & fo/fo/en/(en) & en/en/en/en & -/-/-/-/-/-/-/- & -/-/-/- \\
Z1058 & 13 & 1400 & 30 & -/per/(per)/L/L & fo/fo/L/L & en/en/en/(en) & -/-/-/-/-/-/-/- & -/-/-/- \\
Z1060 & 13 & 1500 & 20 & -/-/-/-/- & fo/L/L/L & en/en/L/L & -/-/-/-/-/-/-/- & -/-/-/- \\
Z1140 & 13 & 1600 & 10 & -/-/-/-/- & -/-/-/- & -/-/-/- & s/s/s/s/s/s/s/(s) & s+en/en/en/en \\
Z1084 & 13 & 1650 & 10 & -/-/-/-/- & -/-/-/- & L/L/-/- & -/-/-/-/-/-/-/- & -/-/-/- \\
Z1207 & 13 & 1700 & 8 & -/-/-/-/- & -/-/-/- & -/-/-/- & s/s/s/s/s/s/(s)/L & s+en/(en)/L/L \\
Z1209 & 13 & 1800 & 5 & -/-/-/-/- & -/-/-/- & -/-/-/- & s/s/s/s/(s?)/L/L/L & s/L/L/L \\
Z1091 & 13 & 1900 & 5 & (per)/-/-/-/- & -/-/-/- & -/-/-/- & -/-/-/L/-/L/L/L & L/L/-/-
\end{tabular}
\end{table}
\end{landscape}

@@

\begin{figure}[ht!]
  \centering
      \includegraphics[width=0.8\textwidth]{figures/forsterite}
      \caption{The forsterite-water phase diagram at 13 GPa, based on current experimental results. The thermodynamics of pure forsterite are taken from the literature \citep{SLB2011}, and the properties of the anhydrous melt are from the mixing model described in the text. At high temperatures, the forsterite liquidus is metastable with respect to periclase + melt.}
  \label{fig:foH}
\end{figure}
\clearpage

\subsubsection{MgSiO$_3$-H$_2$O}
The melting of high pressure clinoenstatite in the presence of water is shown in Figure \ref{fig:eoH}. Unlike the depression of the forsterite or periclase liquidi, the observed depression of melting at low water contents exceeds that predicted by even the most extreme commonly-used ideal mixing model, where all oxygens in the melt are equivalent and available for protonation, and molecular H$_2$O is absent \citep{SS1985}. The liquidus is in excellent agreement with that estimated previously \citep{YII2004}, as is the absence of stishovite in any of the sample chambers. We also note that although we do not have any evidence for melt-fluid immiscibility, the very small observed dT/dX at ca. 1500 $^{\circ}$C results in a very rapid change from water-rich to water-poor fluids with increasing temperature, in agreement with \cite{Inoue1994} and \cite{YII2004}.

\begin{figure}[ht!]
  \centering
      \includegraphics[width=0.8\textwidth]{figures/enstatite}
  \caption{The clinoenstatite-water phase diagram at 13 GPa, based on current experimental results. The thermodynamics of pure high pressure clinoenstatite are taken from the literature \citep{SLB2011}, and the properties of the anhydrous melt are from the mixing model described in the text.}
  \label{fig:eoH}
\end{figure}

@@

\begin{figure}[ht!]
  \centering
      \includegraphics[width=0.8\textwidth]{figures/stishovite}
      \caption{The silica-water phase diagram at 13 GPa, based on current experimental results. The thermodynamics of stishovite and the anhydrous melt are described in the text. At high temperatures, the stishovite liquidus is metastable with respect to coesite and melt.}
  \label{fig:SH}
\end{figure}

@@

\clearpage
\subsection{The MgO-SiO$_2$-H$_2$O system}
The set of experimental analyses can be used to create a preliminary ternary liquidus diagram (Figure \ref{fig:ternary}).

\cite{YII2004} suggested that there was a sharp break in the liquidus at the enstatite-stishovite cotectic (their Figure 6). However, this prediction breaks Schreinemakers Rules (specifically, that no one assemblage can occupy more than 180$^{\circ}$ around an invariant point).

\citep{MSUP2007}

\begin{figure}[ht!]
  \centering
      \includegraphics[width=0.8\textwidth]{figures/experimental-ternary}
  \caption{A preliminary ternary liquidus diagram at 13 GPa, based on current experimental results.}
  \label{fig:ternary}
\end{figure}

\section{Discussion}
\subsection{Melting in the MgO-SiO$_2$-H$_2$O system}
\subsection{Water partitioning in the mantle transition zone}
% Olivine, wadsleyite and ringwoodite

Partitioning of hydrogen between mineral and melt phases can be described using a partition coefficient D:
\begin{equation}
  D^{a/b}_{H_2O} = c^a_{H_2O} / c^b_{H_2O}
\end{equation}
where $c^i_{H_2O}$ is the concentration of H$_2$O (in wt\%) in phase $i$ \citep{KB2006}. In quenchable phases, there are a number of ways that concentration can be measured, including FTIR, SIMS and ERDA. In the case of melts, which are unquenchable, $c^{melt}_{H_2O}$ in partitioning studies has most commonly been estimated by mass balance or by assuming that the deficit in microprobe totals is due to hydrogen. In two studies on Fe-free wadsleyite, \cite{DDFK2005} and \cite{LSOK2011} present estimates of water content in melts calculated from microprobe deficits. A similar study on ringwoodite was undertaken by \cite{OMY2000}. The T-X$_{H_2O}$ dependencies of the melts in equilibrium with wadsleyite or ringwoodite are strikingly different (Figure \ref{fig:partitioning}). For example, melt in equilibrium with wadsleyite at 15 GPa and 1400 $^{\circ}$C was estimated to contain 10.6-13.3 wt\% H$_2$O, while melt compositions in equilibrium with ringwoodite at 20 GPa and 1400 $^{\circ}$C suggest water contents of 57-66 wt\% H$_2$O. Modelled Mg$_2$SiO$_4$ melting temperatures increase by only 40 K between 15 and 20 GPa, and entropies of melting at these two pressures are not very different (92.4 vs. 97 J/K/mol), so unless the thermodynamics of mixing in the liquid change markedly between these pressures, estimates of $c^{melt}_{H_2O}$ for the ringwoodite or wadsleyite studies (or both) must be somewhat inaccurate. We note that two of the three melt compositions in equilibrium with ringwoodite fall above the temperatures indicating that hydrogen exists only as molecular H$_2$O \citep{SS1985}. In addition, the calculated liquidus temperatures in the wadsleyite studies at 15 GPa are lower than those reported at 5.5 GPa \citep{Inoue1994}. Both of these observations appear highly unlikely.

@@

\begin{figure}[ht!]
  \centering
      \includegraphics[width=0.8\textwidth]{figures/wad_rw}
  \caption{Water concentrations in wadsleyite, ringwoodite and melt, modelled at 15 and 20 GPa. Experimental data is taken from three published studies \citep{OMY2000,DDFK2005,LSOK2011}.}
  \label{fig:wad_rw}
\end{figure}

\begin{figure}[ht!]
  \centering
  \includegraphics[width=0.8\textwidth]{figures/wad_rw_partitioning}
  \caption{Modelled partition coefficient between wadsleyite and ringwoodite at 18 GPa.}
  \label{fig:partitioning}
\end{figure}

@@

These models allow us to predict the partition coefficients of water between wadsleyite, ringwoodite and melt at the 520 km discontinuity ($\sim$18 GPa). $D^{wad/ring}_{H_2O}$ increases from $\sim$0.7 at 1000 $^{\circ}$C to almost $\sim$0.9 at 2000 $^{\circ}$C. These values imply that water saturated ringwoodite in the lower mantle will partially melt as it transforms to wadsleyite during upwelling, which could explain observations of low velocity layers at ca. 520 km depth \citep[e.g.][]{JDH2010}. The small temperature dependence implies that repartitioning of water during secular cooling of the Earth is likely to be more minor (and in the opposite direction) to that proposed by \cite{DDFK2005}.

\subsection{Deep water release from subduction zones}
% Densities
\cite{MF2010}
\cite{BK2003}
\cite{MSK2008}

% Viscosities
% Shear viscosity 
% Density

>>>>

\begin{abstract}
Hydrous melting at high pressures affects the physical properties, dynamics and chemical differentiation of the Earth. However, probing the compositions of hydrous melts at the conditions of the mantle transition zone has traditionally been challenging. In this study, we conducted high pressure multianvil experiments at 13 GPa between 1200 and 1900 $^{\circ}$C to investigate the liquidus in the system MgO-SiO$_2$-H$_2$O. Water-rich starting compositions were created using platinic acid (H$_2$Pt(OH)$_6$) as a novel water source. As MgO:SiO$_2$ ratios decrease, the liquidus curve develops an increasingly pronounced concave-down topology. At slab temperatures, the forsterite- and enstatite-water liquidii have similar water contents. Preliminary experiments on the SiO$_2$-H$_2$O binary reveal a pronounced melting point depression up to about 20 mol\% H$_2$O. This melting point depression exceeds estimates from published models, and requires either significant non-ideality between species in the melt, or distinct differences in the ease of protonation of bridging oxygens. We discuss the implications of this work on the migration of deep-seated slab fluids, and the partitioning of water between nominally anhydrous phases and melts.
\end{abstract}

@@

\begin{landscape}
\begin{table}[ht!]
\caption{Experimental run conditions and run products determined by SEM/EDS. Compositions are listed in order of increasing molar H$_2$O content. Unused compositions for each run are marked by an en-dash. Minor solids are listed in brackets. Mineral abbreviations are as follows: br - brucite, per - periclase, co - hydroxychondrodite, en - clinoenstatite, s - stishovite.}
\label{table:experiments}
\begin{tabular}{llllccccc}
Expt \# & P (GPa) & T ($^{\circ}$C) & t (min) & brucite+PtAc & forsterite+PtAcid & enstatite+PtAcid & quartz+PtAcid & br+q \\
Z1063 & 13 & 1200 & 40 & -/br/br/br/L & co/fo/fo/(fo) & en/en/en/en & -/-/-/-/-/-/-/- & -/-/-/- \\
Z1085 & 13 & 1300 & 40 & -/-/per/L/L & -/fo/(en)/L & -/-/-/- & -/-/-/-/-/-/-/- & -/-/-/- \\
Z1079 & 13 & 1300 & 35 & -/per/per/L/L & fo/fo/en/(en) & en/en/en/en & -/-/-/-/-/-/-/- & -/-/-/- \\
Z1058 & 13 & 1400 & 30 & -/per/(per)/L/L & fo/(fo)/L/L & en/en/en/(en) & -/-/-/-/-/-/-/- & -/-/-/- \\
Z1060 & 13 & 1500 & 20 & -/-/-/-/- & (fo)/L/L/L & en/en/L/L & -/-/-/-/-/-/-/- & -/-/-/- \\
Z1140 & 13 & 1600 & 10 & -/-/-/-/- & -/-/-/- & -/-/-/- & s/s/s/s/s/s/s/(s) & s+en/en/en/en \\
Z1084 & 13 & 1650 & 10 & -/-/-/-/- & -/-/-/- & L/L/-/- & -/-/-/-/-/-/-/- & -/-/-/- \\
Z1207 & 13 & 1700 & 8 & -/-/-/-/- & -/-/-/- & -/-/-/- & s/s/s/s/s/s/(s)/L & s+en/(en)/L/L \\
Z1209 & 13 & 1800 & 5 & -/-/-/-/- & -/-/-/- & -/-/-/- & s/s/s/s/(s?)/L/L/L & s/L/L/L \\
Z1091 & 13 & 1900 & 5 & (per)/-/-/-/- & -/-/-/- & -/-/-/- & -/-/-/L/-/L/L/L & L/L/-/-
\end{tabular}
\end{table}
\end{landscape}

@@

\begin{figure}[ht!]
  \centering
      \includegraphics[width=0.8\textwidth]{figures/forsterite}
      \caption{The forsterite-water phase diagram at 13 GPa, based on current experimental results. The thermodynamics of pure forsterite are taken from the literature \citep{SLB2011}, and the properties of the anhydrous melt are from the mixing model described in the text. At high temperatures, the forsterite liquidus is metastable with respect to periclase + melt. The preferred fit through the data is given by $G^{\circ} = 75 \cdot (1420-T)$ J/mol.}
  \label{fig:foH}
\end{figure}
\clearpage

\subsubsection{MgSiO$_3$-H$_2$O}
The melting of high pressure clinoenstatite in the presence of water is shown in Figure \ref{fig:eoH}. Unlike the depression of the forsterite or periclase liquidi, the observed depression of melting at low water contents exceeds that predicted by even the most extreme commonly-used ideal mixing model, where all oxygens in the melt are equivalent and available for protonation, and molecular H$_2$O is absent \citep{SS1985}. The liquidus is in excellent agreement with that estimated by \cite{YII2004}, as is the absence of stishovite in any of the sample chambers. We note that although we do not have any evidence for melt-fluid immiscibility, the very small observed dT/dX at ca. 1500 $^{\circ}$C results in a very rapid change from water-rich to water-poor fluids with increasing temperature, in agreement with \cite{Inoue1994} and \cite{YII2004}.

\begin{figure}[ht!]
  \centering
      \includegraphics[width=0.8\textwidth]{figures/enstatite}
  \caption{The clinoenstatite-water phase diagram at 13 GPa, based on current experimental results. The thermodynamics of pure high pressure clinoenstatite are taken from the literature \citep{SLB2011}, and the properties of the anhydrous melt are from the mixing model described in the text. The line K=K(T) is given by $G^{\circ} = 120 \cdot (1750-T)$ J/mol.}
  \label{fig:eoH}
\end{figure}

@@

\begin{figure}[ht!]
  \centering
      \includegraphics[width=0.8\textwidth]{figures/stishovite}
      \caption{The silica-water phase diagram at 13 GPa, based on current experimental results. The thermodynamics of stishovite and the anhydrous melt are described in the text. At high temperatures, the stishovite liquidus is metastable with respect to coesite and melt. The line K=K(T) is given by $G^{\circ} = 100 \cdot (1200-T)$ J/mol.}
  \label{fig:SH}
\end{figure}

@@

\clearpage
\subsection{The MgO-SiO$_2$-H$_2$O system}
The set of experimental analyses can be used to create a preliminary ternary liquidus diagram (Figure \ref{fig:ternary}). It should be noted that the periclase liquidus field and periclase-forsterite cotectic are poorly constrained by the data. The low temperature extension of the cotectic is chosen to fit the reported Mg:Si ratios of melts in equilibrium with forsterite and various DHMS phases at 13.5 GPa \citep{MSUP2007}.

% Yamada et al. Figure 6 breaks Schreinemakers rules at high water contents... no reason to mention this in text, though...

\begin{figure}[ht!]
  \centering
      \includegraphics[width=0.8\textwidth]{figures/experimental-ternary}
  \caption{A preliminary ternary liquidus diagram at 13 GPa, based on current experimental results. Contours are in $^{\circ}$C, and mineral abbreviations are liquidus fields as follows: per - periclase, fo - forsterite, en - clinoenstatite, stv - stishovite, coe - coesite, phB - phase B, br - brucite, chond - hydroxychondrodite, ice - ice VII.}
  \label{fig:ternary}
\end{figure}

\section{Discussion}
\subsection{Melting in deep subduction zones and atop the mantle transition zone}
Our melt model in the MSH system serves as a useful tool for investigating melting at high pressures in the mantle. In particular, the forsterite-enstatite cotectic implies that hydrous melts in equilibrium with depleted peridotites at typical mantle temperatures of 1500 $^{\circ}$C will contain $\sim$ 48 mol\% H$_2$O and an Mg:Si ratio of about 2 (the same as forsterite), corresponding to about 26 wt\% H$_2$O. Such compositions are far too water-rich to be neutrally buoyant relative to a forsterite + enstatite assemblage \citep{MSK2008}, and will rise through the mantle. Adding further components to the system will modify this behaviour. In particular, the addition of iron lowers anhydrous melting temperatures and increases melt densities. Thus, we cannot directly use our results to confirm or refute the viability of the water-filter model proposed by \cite{BK2003}, which requires neutrally buoyant or dense melts at the 410 km discontinuity. However, the extremely low dT/dX revealed by our experiments implies that between 1300 and 1700 $^{\circ}$C, water contents in the melt drop from 62 mol\% (34 wt\%) to around 30 mol\% (16 wt\%). Such large decreases in water content may lead to an abrupt sign change in density, where melts are buoyant at relatively low temperatures but dense at high temperatures. This would explain the preservation of high pressure phases in diamonds \citep{Pearsonetal2014}, which must have risen quickly from the mantle transition zone within a buoyant melt \citep{Keppler2014}. Large decreases in water contents are also likely to have a profound effect on melt viscosity, and may encourage reaction-infiltration instabilities encouraging melt channelisation \citep{SKA2001}.

% Densities
%\cite{MF2010}
%\cite{BK2003}
%\cite{MSK2008}

\subsection{Water partitioning in the mantle transition zone}
% Olivine, wadsleyite and ringwoodite

Partitioning of hydrogen between mineral and melt phases can be described using a partition coefficient D:
\begin{equation}
  D^{a/b}_{H_2O} = c^a_{H_2O} / c^b_{H_2O}
\end{equation}
where $c^i_{H_2O}$ is the concentration of H$_2$O (in wt\%) in phase $i$ \citep{KB2006}. In quenchable phases, there are a number of ways that concentration can be measured, including FTIR, SIMS and ERDA. In the case of melts, which are unquenchable, $c^{melt}_{H_2O}$ in partitioning studies has most commonly been estimated by mass balance or by assuming that the deficit in microprobe totals is due to hydrogen. In two studies on Fe-free wadsleyite, \cite{DDFK2005} and \cite{LSOK2011} present estimates of water content in melts calculated from microprobe deficits. A similar study on ringwoodite was undertaken by \cite{OMY2000}. The T-X$_{H_2O}$ dependencies of the melts in equilibrium with wadsleyite or ringwoodite are strikingly different (Figure \ref{fig:wad_rw_melt}). For example, melt in equilibrium with wadsleyite at 15 GPa and 1400 $^{\circ}$C was estimated to contain 10.6-13.3 wt\% H$_2$O, while melt compositions in equilibrium with ringwoodite at 20 GPa and 1400 $^{\circ}$C suggest water contents of 57-66 wt\% H$_2$O. Modelled Mg$_2$SiO$_4$ melting temperatures increase by only 40 K between 15 and 20 GPa, and entropies of melting at these two pressures are not very different (92.4 vs. 97 J/K/mol), so unless the thermodynamics of mixing in the liquid change markedly between these pressures, estimates of $c^{melt}_{H_2O}$ for the ringwoodite or wadsleyite studies (or both) must be somewhat inaccurate. We note that two of the three melt compositions in equilibrium with ringwoodite fall above the temperatures indicating that hydrogen exists only as molecular H$_2$O \citep{SS1985}. In addition, the calculated liquidus temperatures in the wadsleyite studies at 15 GPa are lower than those reported at 5.5 GPa \citep{Inoue1994}. Both of these observations appear highly unlikely.

@@

\begin{figure}[ht!]
  \centering
      \includegraphics[width=0.8\textwidth]{figures/wad_rw}
  \caption{Water concentrations in wadsleyite, ringwoodite and melt, modelled at 15 and 20 GPa. Experimental data is taken from three published studies \citep{OMY2000,DDFK2005,LSOK2011}.}
  \label{fig:wad_rw_melt}
\end{figure}

@@

These models allow us to predict the partition coefficients of water between wadsleyite, ringwoodite and melt at the 520 km discontinuity ($\sim$18 GPa; Figure \ref{fig:partitioning}). $D^{wad/ring}_{H_2O}$ increases from $\sim$0.7 at 1000 $^{\circ}$C to almost $\sim$0.9 at 2000 $^{\circ}$C. These values imply that water saturated ringwoodite in the lower mantle will partially melt as it transforms to wadsleyite during upwelling, which could explain observations of low velocity layers at ca. 520 km depth \citep[e.g.][]{JDH2010}. The small temperature dependence implies that repartitioning of water during secular cooling of the Earth is likely to be more minor (and in the opposite direction) to that proposed by \cite{DDFK2005}.

\begin{figure}[ht!]
  \centering
  \includegraphics[width=0.8\textwidth]{figures/wad_rw_partitioning}
  \caption{Modelled partition coefficient between wadsleyite, ringwoodite and melt at 18 GPa. Note the separate scales for solid/solid and solid/melt partitioning.}
  \label{fig:partitioning}
\end{figure}
